Mass and occupation fraction of dark matter halos hosting Lyman-$\alpha$ emitters at $z\sim 3$

### 2013-12-31 paper/ms.tex
new full version

\begin{abstract}
%
We derive constraints on the mass and occupation fraction of dark
matter halos hosting \ly Emitting galaxies (LAEs) at a redshift of
$z=3.1$ by matching the number density and the angular
correlation function between mock and observed fields. We explicitly
take into account the cosmic variance on the typical observed field
size by constructing mock fields from a large cosmological N-body
simulation matching the geometries of observed fields. To populate the
halos in the simulation we use a model where a dark matter halo with
mass in the range $M_{\rm min}<M_{\rm h}<M_{\rm max}$ can only host
one detectable LAE with a probability $f_{\rm occ}$. The main motivation to
build such a model is to avoid the large uncertainties in the
estimation of \ly luminosities. Our main result is that the clustering
and number density information are insufficient to impose a tight
constraint on the {\bf occupation} fraction. On the other hand the
minimum mass and maximum mass can be tightly constrained to the range
$M_{\rm max}<10^{12}$\hMsun and $10^{10}\hMsun\leq M_{\rm min}\leq
10^{11.5}\hMsun$.  We also find that all models have a narrow mass
range $\Delta M \equiv \log_{10}M_{\rm max} - \log_{10} M_{\rm min}$
smaller than $1.0$ dex, with the vast majority in the range $\Delta
M<0.5$ dex. Already published constraints on the mass range and
occupation fraction are a sub-set of the models we report in this
\documentname. Our results present an inconclusive response as to
which is the occupation fraction for LAEs at $z=3.1$. A precise answer
requires observations on larger fields and robust physical models for
the connection between dark matter, star formation and the escape of
\ly radiation. That kind of modelling should also be able to provide a
plausible explanation for the narrow range in mass $\Delta M<1.0$ dex
derived here for the halos hosting LAEs at $z=3.1$.

@@

Figure \ref{fig:restriction_mock} presents the locii of these models
in the parameter space $M_{\rm min}-M_{\rm max}$ and $M_{\rm
  min}-f_{\rm occ}$. With this constraints the number of consistent
models with  $10.5 < \log_{10}M_{\rm min}< 11.0$ are reduced. This 
corresponds to the regions in the parameter space in
Figure \ref{fig:landscape} that already had a low number of 
consistent mock surveys. On the other hand, from the right panel in
Figure \ref{fig:restriction_mock} one can see that the favored
occupation fraction  values are still in a range $0\leq f_{\rm
  occ}\leq 1.0$.

@@

Figure \ref{fig:correlation_parameters} shows the results for the
best estimates of $\theta_{0}$-$\beta$  used in the ACF
parameterization.  Blue circles with error bars represent the results
from the mocks and the horizontal line the obsrvational results of
\cite{Ouchi2010} over a field with average number density where the
$\beta$ parameter was fiexed in the fit. From this Figure we observe
that there are models that can be ruled out as inconsistent with the
angular correlation function.

Figure \ref{fig:restriction_mock_and_f_occ_corr} presents the
remainingn consistent models in the planes $M_{\rm min}$-$M_{\rm
  max}$ and $M_{\rm   min}$-$f_{\rm occ}$. A model is consistent with
observations if there is a $1$-$\sigma$ overlap between both the
correlation length $\theta_0$ and the power $\beta$.  We find that
we end up with $40$ models consistent with the ACF constraints.

@@

\section{Discussion}

@@

{\bf The models in the third family are barely consistent with the
  constraints from the ACF. They have mean values of $\theta_0$
  outside the observational uncertainties, but have
  are considered as consistent because their error bars overlap. These
  models they have another particular feature: their minimum mass is
  exactly $M_{\rm min}=10^{10.9}\hMsun$. Because these models are
  are barely compatible with observations and are a minority of the
  consisten models, we exclude them now from the discussion.}

There are two interesting feature in the remaining two first
families. First, the occupation fraction can take any value from $0.1$
to $1.0$. Second, the halos hosting LAEs all have a narrow mass range
$\Delta M\leq 1.0$.

{\bf The existence of models with high occupation fractions $f_{\rm
    occ}>0.2$ is unexpected from previous clustering analysis that do
  not take  account the effect of cosmic variance in the same way we
  propose in this \documentname. This suggests that taking explicitly
  cosmic variance into account hinders the possibility of constraining the LAEs
  occupation fraction.}

\subsection{Physical interpretations}

{\bf The occupation fraction is midly correlated with a model's
  minimum mass. As can be observed in the right panel of Figure 6, the
  majority of  models with a high occupation fraction have  a minimum
  mass $M_{\rm  min}> 10^{10.7}$\hMsun. Therefore, low values for the
  occupation fraction would be favored if one could rule out models
  with a minimum mass above that threshold.}

{\bf On the other hand, a robust conclusion from the two first families
  is that the possible models have a narrow halo mass range $\Delta M
  <1.0$. How can we give a physical interpretation to the existence of
a narrow mass range? We can start with a reasonable assumption of star
formation rate increases with halo's mass. In this way a narrow
$\Delta M$ implies a cut both at the low and high mass end. }

A cut at the low mass end, $M_{\rm min}$, can be
interpreted in terms of the minimal halo star formation rate needed to
produce the \ly luminosity needed to be above a given detection
threshold.   A cut at higher halo masses $M_{\rm  max}$ requires a
different justification. There are two complementary physical
scenarios that could provide it.

One scenario can be presented in terms of a decreasing escape fraction
of \ly radiation in massive systems. Detailed galaxy formation models
support the idea that massive galaxies with higher metallicities have
larger dust contents and less concentrated ISM than lower mass
systems. Due to the resonant nature of the \ly line the probability of
absorption  of \ly photons increases in massive systems, producing
high absorption of the \ly line but not of UV continuum or other
non-resonant lines \citep{Laursen2009,ForeroRomero2011}. In a second
scenario larger systems have more extended gaseous envelopes which due
to resonance effects of the \ly line, induces a low surface brightness
and a broader line, making these systems less observable in narrow
band filter surveys \citep{Laursen2009,Zheng2010}.

@@

There are two important caveats for the work presented here. The first is the
assumption of a single LAE per dark matter halo. This contradicts the
general expectation of dark matter sub-halos in the simulation to host
satellite galaxies. However, it has been found in analysis based on
the shape of the correlation function \citep{Jose2013b} that satellite
galaxies are not a dominant population, making our initial
approximation a reasonable one. Therefore we conclude that including
the effect of satellite galaxies won't change the main results
reported in this \documentname.

The second caveat are the precise values for the mass intervals. These
values are defined from the halos defined in using the FOF halo
finder. Different halo finders and definitions for the detection
density threshold can yield different masses up to a factor. For
instance a Friends-of-Friends algorithm with linking length $l=0.20$
times the average inter-particle distance finds halos on average $1.4$
times less massive than halos defined  with an spherical overdensity
algorithm halos \citep{Bolshoi}. Therefore, the mass values for
$M_{\rm min}$ and $M_{\rm max}$ should not be considered exact within
less than $\sim 0.2$ dex.

@@

In this \documentname we look for constraints on the preferred mass
and occupation fraction of dark matter halos hosting Lyman Alpha Emitters at
redshift $z=3.1$ in a $\Lambda$CDM cosmology. We perform this study
paying special attention to the impact of cosmic variance on these
results. To this end we build a large number of mock catalogs matching
observational geometries. The mocks are constructed from a N-body simulation
following a simple recipe to assign a single LAE to each halo. Only
a fraction $f_{\rm occ}$ of halos with a mass range  $M_{\rm
  min}<M_{\rm h}<M_{\rm   max}$ can host a LAE. We proceed with a
thorough exploration of the space of free parameters $M_{\rm min},
M_{\rm max}$ and $f_{\rm occ}$ to find mocks that are consistent with
two observational constraints: the surface number density and the
angular correlation function. Out of the initial $9000$ 
combinations of parameters in the model we find $23$ arrangements
consistent with observations.

We find that the observational information we use is insufficient to
impose a tight constraint on the escape fraction. On the other hand
the minimum mass and maximum mass are tightly constrained as follows
$M_{\rm max}<10^{12}$\hMsun and $10^{10}\hMsun\leq M_{\rm min}\leq
10^{11.5}\hMsun$. Furthermore, we find that the mass range defined as
$\Delta M=\log_{10}M_{\rm max} - \log_{10}M_{\rm min}$ is always
$\Delta M\leq 1.0$.

The wide range in solutions is facilitated by the large dispersion in
the statistics derived from the mocks. Nevertheless, all the halo mass
ranges and occupation fractions deduced previous analysis
\citep[i.e.][]{Gawiser2007,Ouchi2010} are a subset of the models we
find in this \documentname. All of our models support the notion that
the most massive halos at $z=3.1$ do not host the brightest LAEs.

The existence of a narrow mass range for halos hosting LAEs, $\Delta
M\leq 1.0$ dex, has two different explanations at least. The
first is having LAEs with a decreasing \ly escape fraction with
increasing mass, the second is having larger screening effects by
neutral Hydrogen around the most massive systems
\citep{Laursen2009,ForeroRomero2011}.

However, there are extreme cases with a very narrow $\Delta M<0.3$, meaning that
there is barely a factor of $\sim 2$ between the minimum and halo
mass. If these models turn out to be confirmed, this would be a great
challenge for galaxy formation models to explain how is that LAEs can
be hosted in such a narrow range of halo mass.

In summary, we show how spatial information of LAEs cannot
put a tight constraint on their host halos and occupation fraction. We
foresee that the new observations with new instruments (such as MUSE,
Hyper SuprimeCam and HETDEX) covering larger fields and a wider range
of luminosities will be key in imposing tighter constraints on the
properties of dark matter halos hosting LAEs. At the same time,
additional modeling for \ly radiation transfer , is needed to put a
tighter constrain on the \ly escape fraction in high redshift
galaxies, also paying attention to newly framed physical phenomena, such as the
stochasticity \citep{ForeroRomero2013} in the star formation process,
which might play a role in inducing detection biases in high redshift LAEs.

>>>>

\begin{abstract}
%
We derive constraints on the mass and occupation fraction of dark
matter halos hosting \ly Emitting galaxies (LAEs) at a redshift of
$z=3.1$ by matching the number density and the angular
correlation function between mock and observed fields. We explicitly
take into account the cosmic variance on the typical observed field
size by constructing mock fields from a large cosmological N-body
simulation matching the geometries of observed fields. To populate the
halos in the simulation we use a model where a dark matter halo with
mass in the range $M_{\rm min}<M_{\rm h}<M_{\rm max}$ can only host
one detectable LAE with a probability $f_{\rm occ}$. The main motivation to
build such a model is to avoid the large uncertainties in the
estimation of \ly luminosities. Our main result is that the clustering
and number density information are insufficient to impose a tight
constraint on the {\bf occupation} fraction. On the other hand the
minimum mass and maximum mass can be tightly constrained to the range
$M_{\rm max}<10^{12}$\hMsun and $10^{10}\hMsun\leq M_{\rm min}\leq
10^{11.5}\hMsun$.  We also find that the consistent models have a narrow mass
range $\Delta M \equiv \log_{10}M_{\rm max} - \log_{10} M_{\rm min}$
smaller than $1.0$ dex, with the vast majority in the range $\Delta
M<0.5$ dex. Already published constraints on the mass range and
occupation fraction are a sub-set of the models we report in this
\documentname. Our results present an inconclusive response as to
which is the occupation fraction for LAEs at $z=3.1$. A precise answer
requires observations on larger fields and robust physical models for
the connection between dark matter, star formation and the escape of
\ly radiation. That kind of modelling should also be able to provide a
plausible explanation for the narrow range in mass $\Delta M<1.0$ dex
derived here for the halos hosting LAEs at $z=3.1$.

@@

Figure \ref{fig:restriction_mock} presents the locii of these models
in the parameter space $M_{\rm min}-M_{\rm max}$ and $M_{\rm
  min}-f_{\rm occ}$. With this constraint the number of consistent
models with  $10.5 < \log_{10}M_{\rm min}< 11.0$ are reduced. This 
corresponds to the regions in the parameter space in
Figure \ref{fig:landscape} that already had a low number of 
consistent mock surveys. On the other hand, from the right panel in
Figure \ref{fig:restriction_mock} we see that the favored occupation
fraction  still remain unconstrained $0\leq f_{\rm  occ}\leq 1.0$.

@@

Figure \ref{fig:correlation_parameters} shows the results for the
best estimates of $\theta_{0}$-$\beta$  used in the ACF
parameterization.  Blue circles with error bars represent the results
from the mocks and the horizontal line the obsrvational results of
\cite{Ouchi2010} over a field with average number density where the
$\beta$ parameter was fixed in the fit. From this Figure we observe
that there are models with too large values of $\theta_0$ and $\beta$
that can be ruled out as inconsistent with the angular correlation
function.

Figure \ref{fig:restriction_mock_and_f_occ_corr} presents the
remaining consistent models in the planes $M_{\rm min}$-$M_{\rm
  max}$ and $M_{\rm   min}$-$f_{\rm occ}$. A model is consistent with
observations if there is a $1$-$\sigma$ overlap between both the
correlation length $\theta_0$ and the exponent $\beta$.  We find that
we end up with $40$ models consistent with the ACF constraints.

@@

\section{Discussion}
\label{sec:discussion}

@@

{\bf The models in the third family are barely consistent with the
  constraints from the ACF. They have mean values of $\theta_0$
  outside the observational uncertainties, but have
  are considered as consistent because their large error barsoverlap. These
  models they have another particular feature: their minimum mass is
  exactly $M_{\rm min}=10^{10.9}\hMsun$. Because these models are
  are barely compatible with observations and are a minority of the
  consistent models, we exclude them from the rest of the discussion.}

\subsection{Two relevant features}
There are two interesting features in the remaining two first
families. First, the occupation fraction can take any value from $0.1$
to $1.0$. Second, the halos hosting LAEs all have a narrow mass range
$\Delta M\leq 1.0$.

{\bf The existence of models with high occupation fractions $f_{\rm
    occ}>0.2$ is unexpected from previous clustering analysis that do
  not take  account the effect of cosmic variance in the same way we
  propose in this \documentname. Nevertheless, the mass range
  predicted by our models is consistent with already published
  observational results. This suggests that taking explicitly cosmic
  variance into account hinders the possibility of constraining the
  LAEs occupation fraction.}

{\bf Concerning the narrow mass range $\Delta M
  <1.0$ the question arises of how can we give a physical
  interpretation for the existence of such  a narrow mass range. We
  can start with a reasonable   assumption. Namely that star formation
  rate increases with halo's  mass.}

Under this assumption the cut at the low mass end, $M_{\rm min}$, can be
interpreted in terms of the minimal star formation rate required to
produce a \ly luminosity above observational detection thresholds.  
A cut at higher halo masses $M_{\rm  max}$ requires a
different justification. There are two complementary physical
scenarios that could provide

One scenario can be presented in terms of a decreasing escape fraction
of \ly radiation in massive systems. Detailed galaxy formation models
support the idea that massive galaxies with higher metallicities have
larger dust contents and a less concentrated ISM than lower mass
systems. Due to the resonant nature of the \ly line the probability of
absorption  of \ly photons increases in massive systems, producing
high absorption of the \ly line but not of UV continuum or other
non-resonant lines \citep{Laursen2009,ForeroRomero2011}. In a second
scenario larger systems have more extended gaseous envelopes which due
to resonance effects of the \ly line, induces a low surface brightness
and a broader line, making these systems less observable in narrow
band filter surveys \citep{Laursen2009,Zheng2010}.

@@

There are two important caveats for the work presented here. The first is the
assumption of a single LAE per dark matter halo. This contradicts the
general expectation of dark matter sub-halos in the simulation to host
satellite galaxies. However, it has been found in analysis based on
the shape of the correlation function \citep{Jose2013b} that satellite
galaxies are not a dominant population, making our initial
approximation a reasonable one. Therefore we conclude that including
the effect of satellite galaxies does not change the main results
reported in this \documentname.

The second caveat are the precise values for the mass intervals. These
values are quoted from halos defined using a FOF halo
finder. Different halo finders and definitions for the detection
density threshold can yield different masses up to a factor $\sim 2$\citep{More2011}. For 
instance a Friends-of-Friends algorithm with linking length $l=0.20$
times the average inter-particle distance finds halos on average $1.4$
times less massive than halos defined  with an spherical overdensity
algorithm halos \citep{Bolshoi}. Therefore, the mass values for
$M_{\rm min}$ and $M_{\rm max}$ should not be considered exact within
less than $\sim 0.2$ dex.

@@

In this \documentname we look for constraints on the preferred mass
and occupation fraction of dark matter halos hosting Lyman Alpha Emitters at
redshift $z=3.1$ in a $\Lambda$CDM cosmology. We perform this study
paying special attention to the impact of cosmic variance on these
results. To this end we build a large number of mock catalogs matching
observational geometries. The mocks are constructed from a N-body simulation
following a simple recipe to assign a single LAE to each halo. Only
a fraction $f_{\rm occ}$ of halos within a mass range  $M_{\rm
  min}<M_{\rm h}<M_{\rm   max}$ can host a LAE.

We then proceed with a
thorough exploration of the space of free parameters $M_{\rm min},
M_{\rm max}$ and $f_{\rm occ}$ to find mocks that are consistent with
two observational constraints: the surface number density and the
angular correlation function. Out of the initial $9000$ 
combinations of parameters in the model we find $40$ combination of parameters
consistent with observations.

We find that {\bf the information for the number density constrains
  the minium mass to the range
  $10^{10}\hMsun < M_{\rm min} < 10^{11.5}\hMsun$ while the clustering
  constrains the upper mass $M_{\rm    max} < 10^{12}\hMsun$. However,
  this is insufficient to impose a tight constraint  on the escape
  fraction.} Furthermore, we find that the mass range defined as
$\Delta M=\log_{10}M_{\rm max} - \log_{10}M_{\rm min}$ is always
$\Delta M\leq 1.0$.

The wide range in values for the escape fraction is allowed by the
large dispersion in the statistics derived from the
mocks. This uncertainty is unexpected from previous studies that
unanimously favor low values for the occupation
fractions \citep[i.e.][]{Gawiser2007,Ouchi2010}. Nevertheless, all the
halo mass ranges and occupation fractions deduced previous analysis
are a subset of the models we find in this \documentname.

All of our preferred models support the notion that the most massive
halos at $z=3.1$ do not host detectable LAEs. The existence of a
narrow mass range for halos hosting LAEs, $\Delta  M\leq 1.0$ dex, has
two different explanations at least. The first is having LAEs with a
decreasing \ly escape fraction with increasing mass, the second is
having larger screening effects by neutral Hydrogen around the most
massive systems \citep{Laursen2009,ForeroRomero2011}.

We also find extreme cases with a very narrow $\Delta M<0.3$, meaning that
there is barely a factor of $\sim 2$ between the minimum and halo
mass. If these models can be confirmed, it would mean a great
challenge for galaxy formation models to explain how is that LAEs can
be hosted in such a narrow range of halo mass.

In summary, we show that {\bf the currently available spatial information of LAEs cannot
put a tight constraint on their occupation fraction of dark matter
halos}. This information can confidently be used to stablish that
these galaxies are hosted in low mass halos with masses within a range
spanning less than a factor of ten. We foresee that the new
observations with new instruments (such as MUSE, Hyper SuprimeCam and
HETDEX) covering larger fields and a wider range of luminosities will
be key in reducing cosmic variance and imposing tighter constraints on
the properties of dark matter halos hosting LAEs. At the same time,
additional modeling for \ly radiation transfer is needed to put a
tighter constrain on the \ly escape fraction in high redshift
galaxies, also paying attention to newly framed physical phenomena,
such as the stochasticity \citep{ForeroRomero2013} in the star
formation process, which might play a role in inducing detection
biases in high redshift LAEs.
